Simple repression draft

### 2020-05-27 doc/section_04_bayesian_inference.tex
MRM pass through the entire Bayesian inference section. Still is missing last part.

\subsection{Transcription factor kinetics can be inferred from FISH measurements}
\label{sec:rep_kinetics_inference}

Now that we have a satisfactory model in hand for constitutive
promoters, we would like to return to the main thread: can we
infer repressor binding and unbinding rates from mRNA distributions
over a population of cells as measured by FISH in~\cite{Jones2014}?
If so, how do these inferred rates compare to direct
single-molecule measurements such as from~\cite{Hammar2014} and
to binding energies such as from~\cite{Garcia2011a}
and~\cite{Razo-Mejia2018}, which were inferred under the assumptions
of the equilibrium models in Figure~\ref{fig1:means_cartoons}(B)?
And can this comparison shed light on the unreasonable effectiveness
of the equilibrium models?

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/main/fig04.pdf}
\caption{\textbf{Simple repression parameter inference and comparison.}
(A) Contours which enclose 50\% and 95\% of the posterior
probability mass are shown for each of several 2D slices of
the 9D posterior distribution. The model assumes one
unbinding rate for each operator (O1, O2, O3) and one binding rate for each aTc
concentration (corresponding to an unknown mean repressor copy number).
(B, upper panel)
Ratios of our inferred unbinding rates are compared with
operator binding energy differences measured by
Garcia and Phillips~\cite{Garcia2011a} (triangles) and
Razo-Mejia et.\ al.~\cite{Razo-Mejia2018} (circles).
Blue glyphs compare O2-O1, while orange compare O1-Oid.
(B, lower panel)
Unbinding rates for O1 (cyan) and Oid (red) inferred in this work are compared with
single-molecule measurements of the same from Hammar et.\ at.\ 2014.
The comparison requires an assumption about mRNA lifetime
$\gamma^{-1}$, which was not measured, and we plot representative
choices assuming $\gamma^{-1}=3$~min (triangles) or $\gamma^{-1}=5$~min (squares).
(C) Posterior predictive checks are shown for each of the
datasets used in the inference of the model in (A). Observed FISH
data from~\cite{Jones2014} is plotted as black lines. The median
posterior predictive ECDF for each condition is plotted as a dark
colored line. Lighter colored bands enclose 95\% of all posterior
predictive samples for a given operator/repressor copy number pair.
Samples are sorted by operator and plotted separately for visual clarity.
The unrepressed promoter, UV5, is shown with each as a reference point.
}
\label{fig4:repressed_post_full}
\end{figure}

As we found in Section~\ref{sec:beyond_means}, for our purposes
the ``right'' model of a constitutive promoter is the bursty
picture, model five in Figure~\ref{fig2:constit_cartoons}.
Therefore our starting point here is the analogous model with
repressor added, model 5 in Figure~\ref{fig1:means_cartoons}.
For a given repressor binding site and copy number,
this model has four rate parameters to be inferred:
the repressor binding and unbinding rates $k_R^+$, and $k_R^-$,
the initiation rate of bursts, $k_i$, and the mean burst size $b$
(we nondimensionalize all of these by the mRNA degradation rate $\gamma$).

Before showing the mathematical formulation of our statistical
inference model, we would like to sketch the intuitive structure.
The dataset from~\cite{Jones2014} we consider consists of
single-cell mRNA FISH measurements of nine different conditions,
spanning several combinations of three unique binding sites and
four unique repressor copy numbers. We assume that the values of
$k_i$ and $b$ are known, since we have already cleanly inferred
them from constitutive promoter data, and further we assume that
these values are the same across datasets with different
repressor binding sites and copy numbers. We assume that there is
one unbinding rate parameter for each repressor binding site,
which is shared across conditions, and likewise one binding rate
for each unique repressor copy number. This makes our model seven
dimensional, or nine if one counts $k_i$ and $b$ as well.

Formally now, denote the set of seven repressor rates to be inferred as
\begin{equation}
\vect{k} =\{k_{Oid}^-, k_{O1}^-, k_{O2}^-,
k_{0.5}^+, k_{1}^+, k_{2}^+, k_{10}^+\}.
\end{equation}
Note the repressor copy numbers are not known directly, so we label their
association rates by the concentration of a small molecule inducer aTc
which indirectly controls the transcription of $\textit{lacI}$.
Also note that the authors of~\cite{Jones2014} report estimates
of LacI copy number per cell rather than direct measurements.
However, these estimates were made \textit{assuming} the validity
of the equilibrium models in Figure~\ref{fig1:means_cartoons},
and since \textit{testing} these models is our goal, we will make
no assumptions about the LacI copy number for given aTc concentrations.

Bayes theorem reads simply
\begin{equation}
p(\vect{k}, k_i, b \mid D)
\propto
p(D \mid\vect{k}, k_i, b) p(\vect{k}, k_i, b),
\end{equation}
where $D$ is the set of all $N$ observed single-cell mRNA counts
across the various conditions. We assume that individual single-cell
measurements are independent so that the likelihood factorizes as
\begin{equation}
p(D \mid\vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid \vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid k_j^+, k_j^-, k_i, b)
\end{equation}
where $k_j^\pm$ represent the appropriate binding and unbinding
rates for the $j$-th measured cell. The probability
$p(m\mid k_j^+, k_j^-, k_i, b)$ appearing in the last expression
is exactly Eq.~\ref{eq:p_m_bursty+rep_appdx}, the steady-state
distribution for our bursty model with repression derived in
Section~\ref{sec:gen_fcn_appdx}, which for completeness we reproduce here as
\begin{equation}
\begin{split}
p(m \mid k_R^+, k_R^-, k_i, b) = & ~\frac{
        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
        }
        {
        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
        }
\frac{b^m}{m!}
\\
&\times {_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b).
\end{split}
\label{eq:p_m_bursty+rep}
\end{equation}
where $\alpha$ and $\beta$, defined for notational convenience, are
\begin{align}
\begin{split}
\alpha &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
\\
\beta &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
\end{split}
\end{align}

@@

Next we specify priors. As for the constitutive model, weakly
informative lognormal priors are a natural choice for all our
rates. We found that if the priors were too weak, our MCMC
sampler would often become stuck in regions of parameter space
with very low probability density, unable to move. We struck a
balance in choosing our prior widths between helping the sampler
run while simultaneously verifying that the marginal posteriors
for each parameter were not artificially constrained or distorted
by the presence of the prior. The only exception to this is the
highly informative priors we placed on $k_i$ and $b$, since we
have strong knowledge of them from our inference of constitutive
promoters above.

With priors and likelihood specified we may write down our complete generative model as
\begin{equation}
\begin{split}
\log_{10}k_i &\sim \text{Normal}(0.725, 0.025)\\
\log_{10}b   &\sim \text{Normal}(0.55, 0.025)\\
\log_{10}k_{0.5}^+ &\sim \text{Normal}(-0.45, 0.3)\\
\log_{10}k_{1}^+   &\sim \text{Normal}(0.6, 0.3)\\
\log_{10}k_{2}^+   &\sim \text{Normal}(1.15, 0.3)\\
\log_{10}k_{10}^+  &\sim \text{Normal}(1.5, 0.3)\\
\log_{10}k_{Oid}^- &\sim \text{Normal}(-0.25, 0.3)\\
\log_{10}k_{O1}^-  &\sim \text{Normal}(0.1, 0.3)\\
\log_{10}k_{O2}^-  &\sim \text{Normal}(0.45, 0.3)\\
m &\sim \text{Likelihood}(k_R^+, k_R^-, k_i, b),
\end{split}
\end{equation}
where the likelihood is specified by Eq.~\ref{eq:p_m_bursty+rep}.
\mmnote{Now that I've typed this out, this notation seems totally
pointless here, especially when the likelihood is not a standard
distribution. Maybe I should make a table or something to record
the prior values, and maybe even that should go to SI?}
We ran MCMC sampling on the full nine dimensional posterior
specified by this generative model. To attempt to visualize this
object, in Figure~\ref{fig4:repressed_post_full}(A) we plot
several two-dimensional slices as contour plots, analogous to
Figure~\ref{fig:constit_post_full}(C). Each of these nine slices
corresponds to the $(k_R^+, k_R^-)$ pair of rates for one of the
conditions from the dataset used to fit the model and gives a
sense of the uncertainty and correlations in the posterior.

>>>>

\subsection{Transcription factor kinetics can be inferred from FISH
measurements}\label{sec:rep_kinetics_inference}

Now that we have a satisfactory model in hand for constitutive promoters, we
would like to return to the main thread: can we infer repressor binding and
unbinding rates from mRNA distributions over a population of cells as measured
by single-molecule FISH in~\cite{Jones2014}? If so, how do these inferred rates
compare to direct single-molecule measurements such as from~\cite{Hammar2014}
and to binding energies such as from~\cite{Garcia2011a}
and~\cite{Razo-Mejia2018}, which were inferred under the assumptions of the
equilibrium models in Figure~\ref{fig1:means_cartoons}(B)? And can this
comparison shed light on the unreasonable effectiveness of the equilibrium
models?

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/main/fig04.pdf}
\caption{\textbf{Simple repression parameter inference and comparison.}
(A) Contours which enclose 50\% and 95\% of the posterior probability mass are
shown for each of several 2D slices of the 9D posterior distribution. The model
assumes one unbinding rate for each operator (Oid, O1, O2) and one binding rate
for each aTc induction concentration (corresponding to an unknown mean repressor
copy number). (B, upper panel) Ratios of our inferred unbinding rates are
compared with operator binding energy differences measured by Garcia and
Phillips~\cite{Garcia2011a} (triangles) and Razo-Mejia et.\
al.~\cite{Razo-Mejia2018} (circles). Blue glyphs compare O2-O1, while orange
compare O1-Oid. (B, lower panel) Unbinding rates for O1 (cyan) and Oid (red)
inferred in this work are compared with single-molecule measurements of the same
from Hammar et.\ at.\ 2014. The comparison requires an assumption about mRNA
lifetime $\gamma^{-1}$, which was not measured, and we plot representative
choices assuming $\gamma^{-1}=3$~min (triangles) or $\gamma^{-1}=5$~min
(squares). (C) Theory-experiment comparison are shown for each of the datasets
used in the inference of the model in (A). Observed single-molecule mRNA counts
data from~\cite{Jones2014} is plotted as black lines. The median of the randomly
generated samples for each condition is plotted as a dark colored line. Lighter
colored bands enclose 95\% of all samples for a given operator/repressor copy
number pair. Samples are sorted by operator and plotted separately for visual
clarity. The unregulated promoter, UV5, is shown with each as a reference
point.}
\label{fig4:repressed_post_full}
\end{figure}

As we found in Section~\ref{sec:beyond_means}, for our purposes the ``right''
model of a constitutive promoter is the bursty picture, model five in
Figure~\ref{fig2:constit_cartoons}. Therefore our starting point here is the
analogous model with repressor added, model 5 in
Figure~\ref{fig1:means_cartoons}(C). For a given repressor binding site and copy
number, this model has four rate parameters to be inferred: the repressor
binding and unbinding rates $k_R^+$, and $k_R^-$, the initiation rate of bursts,
$k_i$, and the mean burst size $b$ (we nondimensionalize all of these by the
mRNA degradation rate $\gamma$).

Before showing the mathematical formulation of our statistical inference model,
we would like to sketch the intuitive structure. The dataset
from~\cite{Jones2014} we consider consists of single-cell mRNA FISH measurements
of nine different conditions, spanning several combinations of three unique
repressor binding sites and four unique repressor copy numbers. We assume that
the values of $k_i$ and $b$ are known, since we have already cleanly inferred
them from constitutive promoter data, and further we assume that these values
are the same across datasets with different repressor binding sites and copy
numbers. In other words, we assume that the regulation of the transcription
factor does not affect the mean burst size nor the burst initiation rate. The
regulation happens as the promoter is taken away from the transcriptionally
active state when the promoter is bound. We assume that there is one unbinding
rate parameter for each repressor binding site, which is shared across aTc
inducer conditions, and likewise one binding rate for each unique repressor copy
number. This makes our model nine dimensional, or nine if one counts $k_i$ and
$b$ as well.

Formally now, denote the set of seven repressor rates to be inferred as
\begin{equation}
\vect{k} =\{k_{Oid}^-, k_{O1}^-, k_{O2}^-,
k_{0.5}^+, k_{1}^+, k_{2}^+, k_{10}^+\},
\end{equation}
where subscripts for dissociation rates $k^-$  indicate the different repressor
binding sites, and subscripts for association rates $k^+$ indicate the aTc
concentration at which the repressor itself was induced. We emphasize that for
this particular dataset the repressor copy numbers are not known directly, so we
label their association rates by the concentration of a small molecule inducer
aTc which indirectly controls the transcription of $\textit{lacI}$. Also note
that the authors of~\cite{Jones2014} report estimates of LacI copy number per
cell rather than direct measurements. However, these estimates were made
\textit{assuming} the validity of the equilibrium models in
Figure~\ref{fig1:means_cartoons}, and since \textit{testing} these models is our
goal, we will make no assumptions about the LacI copy number for given aTc
concentrations.

Having stated the problem, Bayes theorem reads simply
\begin{equation}
p(\vect{k}, k_i, b \mid D)
\propto
p(D \mid\vect{k}, k_i, b) p(\vect{k}, k_i, b),
\end{equation}
where $D$ is again the set of all $N$ observed single-cell mRNA counts
across the various conditions. We assume that individual single-cell
measurements are independent so that the likelihood factorizes as
\begin{equation}
p(D \mid\vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid \vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid k_j^+, k_j^-, k_i, b)
\end{equation}
where $k_j^\pm$ represent the appropriate binding and unbinding
rates out of $\vect{k}$ for the $j$-th measured cell. The probability
$p(m\mid k_j^+, k_j^-, k_i, b)$ appearing in the last expression
is exactly Eq.~\ref{eq:p_m_bursty+rep_appdx}, the steady-state
distribution for our bursty model with repression derived in
Section~\ref{sec:gen_fcn_appdx}, which for completeness we reproduce here as
\begin{equation}
\begin{split}
p(m \mid k_R^+, k_R^-, k_i, b) = & ~\frac{
        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
        }
        {
        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
        }
\frac{b^m}{m!}
\\
&\times {_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b).
\end{split}
\label{eq:p_m_bursty+rep}
\end{equation}
where $\alpha$ and $\beta$, defined for notational convenience, are
\begin{align}
\begin{split}
\alpha &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
\\
\beta &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
\end{split}
\end{align}

@@

Next we specify priors. As for the constitutive model, weakly informative
lognormal priors are a natural choice for all our rates. We found that if the
priors were too weak, our MCMC sampler would often become stuck in regions of
parameter space with very low probability density, unable to move. We struck a
balance in choosing our prior widths between helping the sampler run while
simultaneously verifying that the marginal posteriors for each parameter were
not artificially constrained or distorted by the presence of the prior. The only
exception to this is the highly informative priors we placed on $k_i$ and $b$,
since we have strong knowledge of them from our inference of constitutive
promoters above. All details for our prior distributions are listed in 
Appendix~\ref{sec:bayes}.

We ran MCMC sampling on the full nine dimensional posterior specified by this
generative model. To attempt to visualize this object, in
Figure~\ref{fig4:repressed_post_full}(A) we plot several two-dimensional slices
as contour plots, analogous to Figure~\ref{fig:constit_post_full}(C). Each of
these nine slices corresponds to the $(k_R^+, k_R^-)$ pair of rates for one of
the conditions from the dataset used to fit the model and gives a
sense of the uncertainty and correlations in the posterior.
\mrm{before invoking part (C) of the figure we are missing th explanation for 
part (B). I don't fully understand how these computations were made, so I'll 
let Jupiter detail this.}

% With priors and likelihood specified we may write down our complete generative model as
% \begin{equation}
% \begin{split}
% \log_{10}k_i &\sim \text{Normal}(0.725, 0.025)\\
% \log_{10}b   &\sim \text{Normal}(0.55, 0.025)\\
% \log_{10}k_{0.5}^+ &\sim \text{Normal}(-0.45, 0.3)\\
% \log_{10}k_{1}^+   &\sim \text{Normal}(0.6, 0.3)\\
% \log_{10}k_{2}^+   &\sim \text{Normal}(1.15, 0.3)\\
% \log_{10}k_{10}^+  &\sim \text{Normal}(1.5, 0.3)\\
% \log_{10}k_{Oid}^- &\sim \text{Normal}(-0.25, 0.3)\\
% \log_{10}k_{O1}^-  &\sim \text{Normal}(0.1, 0.3)\\
% \log_{10}k_{O2}^-  &\sim \text{Normal}(0.45, 0.3)\\
% m &\sim \text{Likelihood}(k_R^+, k_R^-, k_i, b),
% \end{split}
% \end{equation}
% where the likelihood is specified by Eq.~\ref{eq:p_m_bursty+rep}.
% \mmnote{Now that I've typed this out, this notation seems totally
% pointless here, especially when the likelihood is not a standard
% distribution. Maybe I should make a table or something to record
% the prior values, and maybe even that should go to SI?}
